NumPy---Array Computation for Python

### 2020-01-08 body.tex
More work outline

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Hugunin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)

@@

% S TODO: the first sentence below could flow a bit better
Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily and, around 2000, wrote a re-implementation of much of
Numeric, called NumArray, to support their need to work on large memory-mapped
arrays and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.

Over the next 10 years, NumPy attracted more scientist developers.
The project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy. At first, NumPy had rather sparse documentation, but the project
developed a new standard for ``docstrings''---text describing each NumPy
function---and used this standard to write comprehensive, high-quality
documentation, integrated with the source code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}. The NumPy documentation format
quickly became a standard for the larger Python community.

@@

\section*{Enabling factors}

@@

As Python's popularity for scientific computing, data science, and machine
learning increased, many new use cases arose.
At the same time, the massive number of scientific Python packages that
built on NumPy meant that it had an unusually high need for stability.

In 2018, we hired two (and later four) full-time developers.
This enabled us to implement several bigger features,
which required a sizeable amount of time, conversations with multiple
stakeholders, as well as longer term focus, and were therefore difficult to
accomplish by volunteer effort alone.
To ensure that this paid effort did not disrupt our community owned character
we formalized the feature proposal process, and constructed a development
roadmap with extensive input and feedback from the community.   In addition,
we hired people who were already active in NumPy and its surrounding communities:
they already shared the project's norms and values.
Funded developers also regularly reviewed other people's code as a part of
their job, thereby supporting the work done by volunteers.

Much of the recent work falls into two categories: improving the library's
interaction with external packages, and, with the benefit of several years'
worth of experience, redesigning core parts of the library.

@@

A second example is the rewrite of NumPy's random number generation facilities.
These have been maintained but unchanged for many years to preserve bit-by-bit
compatibility of random number streams.  A rewrite that allowed for the
underlying pseudorandom algorithms to be exchanged and extended was completed
by a volunteer contributor and integrated by a paid developer.

An example of ongoing work is a redesign of the data type system.  Data types
(dtypes for short) tell us more about the elements stored in a NumPy array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
datetimes, or Python objects, each identified by its own dtype.

@@

\fixme{To be written.}

>>>>

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT.
Hugunin based his package on previous work by Jim Fulton,
then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)

@@

% S TODO: the first sentence below could flow a bit better
Around 200, the Space Telescope Science Institute (STScI) software group wrote
a re-implementation of much of Numeric, called NumArray, to support their
need to work on large memory-mapped arrays and arrays of mixed data type
records \cite{STScI-slither}.
This briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.

Over the next 10 years, NumPy attracted more scientist developers.
The project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy.
The project has comprehensive, high-quality documentation,
integrated with the source
code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

@@

As Python's popularity for scientific computing, data science, and machine
learning increased, many new use cases arose.
At the same time, the massive number of scientific Python packages that
built on NumPy meant that it had an unusually high need for stability.
So to guide our development we formalized the feature proposal process, and
constructed a development roadmap with extensive input and feedback from the
community.

In 2018, we hired two (and later four) full-time developers.
This enabled us, with the benefit of several years' worth of experience,
to redesign core parts of the library.

One example of redesigning core parts of the library is the rewrite of NumPy's
random number generation facilities.
These have been maintained but unchanged for many years to preserve bit-by-bit
compatibility of random number streams.  A rewrite that allowed for the
underlying pseudorandom algorithms to be exchanged and extended was completed
by a volunteer contributor and integrated by a paid developer.

Another ongoing example is a redesign of the data type system.  Data types
(dtypes for short) tell us more about the elements stored in a NumPy array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
datetimes, or Python objects, each identified by its own dtype.

\section*{Array protocol ??}

We are increasingly focused on improving the library's interaction with external
packages.

@@

%"""
%NumPy’s high level ndarray API has been implemented several times outside of
%NumPy itself for different architectures, such as for GPU arrays (CuPy), Sparse
%arrays (scipy.sparse, pydata/sparse) and parallel arrays (Dask array) as well
%as various NumPy-like implementations in the deep learning frameworks, like
%TensorFlow and PyTorch.
%
%Similarly there are many projects that build on top of the NumPy API for
%labeled and indexed arrays (XArray), automatic differentiation (Autograd,
%Tangent), masked arrays (numpy.ma), physical units (astropy.units, pint, unyt),
%etc. that add additional functionality on top of the NumPy API. Most of these
%project also implement a close variation of NumPy’s level high API.
%
%We would like to be able to use these libraries together, for example we would
%like to be able to place a CuPy array within XArray, or perform automatic
%differentiation on Dask array code. This would be easier to accomplish if code
%written for NumPy ndarrays could also be used by other NumPy-like projects.
%"""

@@
